---
title: "hw4"
author: "Jess Kaminsky"
date: "March 26, 2018"
output: html_document
---

```{r setup, include=FALSE, warning=FALSE}
library(AER)

risky <- read.csv("risky_behaviors.csv")
attach(risky)

risky <- risky[-1]
risky$fupacts <- round(risky$fupacts)
risky$couples <- factor(risky$couples)
risky$women_alone <- factor(risky$women_alone)
```

## Question 1

1. Poisson regression: the folderrisky.behaviorcontains data from a random-ized trial targeting couples at high risk of HIV infection. The intervention pro-vided counseling sessions regarding practices that could reduce their likelihoodof contracting HIV. Couples were randomized either to a control group, a groupin which just the woman participated, or a group in which both members ofthe couple participated. One of the outcomes examined after three months was“number of unprotected sex acts.”

# Codebook for Risky Behaviors

# sex: woman/man
# couples: 1 if randomized to couples group; 0 if randomized to other groups
# women_alone: 1 if randomized to women alone group; 0 if randomized to other groups
# bs_hiv: baseline HIV status (negative/positive)
# bupacts: Number of unprotected sex acts at baseline
# fupacts: Number of unprotected sex acts at followup
# Note that fupacts is the outcome and bupacts is a covariate


(a)  Model this outcome as a function of treatment assignment using a Poisson regression. Does the model fit well? Is there evidence of overdispersion?

The following model summary shows the results of fitting a generalized linear model with poisson regression modeling the number of unprotected sex acts at followup as a function of treatment group. Here we are using women_alone and couples as indicator variables for two of the three treatment groups - if both indicator variables are equal to 0 the subject is in the control group where neither partner has received counseling.

```{r, warning=FALSE}
model_a <- glm(fupacts ~ couples + women_alone, family = poisson, data = risky)
summary(model_a)
```
When performing a hypothesis test to test the following hypotheses, where k is the dispersion parameter:
$H_0: k = 1 $
$H_1: k > 1 $
We obtain the following results. With a z-score of 4.7542 and p-value = 9.961e-07, there is strong evidence to conclude that the true dispersion parameter is not equal to 1 - it is significantly greater than 1. There is evidence of overdispersion in this model.
```{r}
dispersiontest(model_a, alternative = "greater")
```

```{r}
## get interpretation for this - should provide evidence of model fitting or not
pchisq(model_a$deviance, df = model_a$df.residual, lower.tail = FALSE)
```

We have further evidence of the model not fitting well from looking at a plot of the residual values against treatment group. Where 0 represents the control group, 1 represents the women alone group, and 2 represents the couples group. These residuals are approximately evenly distributed around 0; however the residuals range from -6.6 to 27.15 indicating a large difference between observed and expected outcome values - including more predictors in the model may decrease this error.
```{r}
plot(couples*2+(women_alone), resid(model_a), xlab = "Treatment Group", ylab = "Residual", main = "Plot of Residuals by Treatment Group")
abline(h=0)
```

(b)  Next extend the model to include pre-treatment measures of the outcome and the additional pre-treatment variables included in the dataset. Does the modelfit well? Is there evidence of overdispersion?

We will now fit the same model as above, but include the other predictors included in the dataset - sex, baseline HIV status, and number of unprotected sex acts at baseline. The model generated is presented below.
```{r}
model_b <- glm(fupacts ~ ., family = poisson, data = risky)
summary(model_b)
```
 
In order to test for overdispersion in this model, we will test the following hypotheses:
$H_0: k = 1 $
$H_1: k > 1 $
We obtain the following results. With a z-score of 5.5689 and p-value = 1.282e-08, there is strong evidence to conclude that the true dispersion parameter is not equal to 1 - it is significantly greater than 1. There is evern stronger evidence of overdispersion in this model compared to the previous model with only treatment group being used to predict the outcome.
```{r}
dispersiontest(model_b, alternative = "greater")
```

```{r}
## get interpretation for this - should provide evidence of model fitting or not
pchisq(model_b$deviance, df = model_a$df.residual, lower.tail = FALSE)
```

We again have further evidence of the model not fitting well from looking at a plot of the residual values against treatment group. The range of residual values here is larger; however, the residual deviance for this model is 10200 indicating a slightly better fit that the previous, simpler model where the residual deviance was 12925.
```{r}
plot(couples*2+(women_alone), resid(model_b), xlab = "Treatment Group", ylab = "Residual", main = "Plot of Residuals by Treatment Group")
abline(h=0)
```

(c)  Fit an overdispersed Poisson model. What do you conclude regarding effectiveness of the intervention?
```{r}
model_a2 <- glm(fupacts ~ couples + women_alone, family = quasipoisson, data = risky)
model_b2 <- glm(fupacts ~ ., family = quasipoisson, data = risky)
summary(model_a2)
summary(model_b2)


pchisq(model_a2$deviance, df = model_a$df.residual, lower.tail = FALSE)
pchisq(model_b2$deviance, df = model_a$df.residual, lower.tail = FALSE)

```

(d)  These data include responses from both men and women from the partici-pating couples. Does this give you any concern with regard to our modelingassumptions?
```{r}
# Yes - they are not independent. 
```

## Question 2

## Question 3
Write an iteratively reweighted least squares algorithm (see slides 8 and 9 of the generalized linear models lecture) to fit Poisson regression. You will need to use your answer to problem 2 to determine what a, b and θ  are. Use the canonical link function for the Poisson which is  g ( μ ) = log a ( μ ) .) θ = h ( μ ) . Test it out on problem 1. How close do you get to the right answer?

```{r}
irls <- function(model, data_vector, beta = 1) {
  beta_val <- model$coefficients[beta + 1]
  n <- data_vector*beta_val
  mu <- e^(n)

  
  
}


irls =
function(A, b, family=binomial, maxit=25, tol=1e-08)
{
  x = rep(0,ncol(A))
  for(j in 1:maxit)
  {
    eta    = A %*% x
    g      = family()$linkinv(eta)
    gprime = family()$mu.eta(eta)
    z      = eta + (b - g) / gprime
    W      = as.vector(gprime^2 / family()$variance(g))
    xold   = x
    x      = solve(crossprod(A,W*A), crossprod(A,W*z), tol=2*.Machine$double.eps)
    if(sqrt(crossprod(x-xold)) < tol) break
  }
  list(coefficients=x,iterations=j)
}
```